Informative Autonomous Sampling of Oceanographic Variables Using Joint Excursion Sets

### 2019-10-08 main_new.tex
updates

The covariance is assumed to be stationary and separable for the two variables. 
The $2 \times 2$ pointwise covariance matrix is set to:
\begin{equation}\label{v0}
\bSigma(\bx)=\mbox{Var} 
\begin{bmatrix}
    \xi_T(\bx) \\
    \xi_S(\bx) 
    \end{bmatrix}=
\begin{bmatrix}
0.25^2 & 0.6 \cdot 0.25^2 \\
0.6 \cdot 0.25^2 & 0.25^2
\end{bmatrix}.
\end{equation}
The spatial correlation is of a Matern type:
\begin{equation}\label{v}
\small
\mbox{Corr} 
\left(
\begin{bmatrix}
    \xi_T(\bx) \\
    \xi_S(\bx) 
    \end{bmatrix},
    \begin{bmatrix}
    \xi_T(\bx') \\
    \xi_S(\bx') 
    \end{bmatrix}
    \right)
    = \begin{bmatrix}
1 & 0.6  \\
0.6  & 1
\end{bmatrix}(1+\phi h)\exp (-\phi h),
\end{equation}
where $h=\sqrt{||\bx-\bx'||^2}$ and $\phi=0.3$ indicates an effective correlation range of about $1200$ m. 
%Figure \ref{fig:stat_design} shows the contour lines for the EP for the reference. We notice the trend of increasing salinity and temperature to the west.  
One realization of the random fields representing salinity and
temperature is shown in Fig. \ref{fig:true_temp} and
\ref{fig:true_sal}. The true ES from this prior realization is shown
in Fig. \ref{fig:ESet}.

@@

We compare three different static designs denoted
\textit{static\_north}, \textit{static\_east}, and
\textit{static\_zigzag} with the three sequential approaches
\textit{naive}, \textit{myopic}, and \textit{look-ahead}. Recall that the static
sampling paths are pre-determined and cannot be changed on the basis
of data. They are representing the pre-planned observation
strategies used in current-day AUV operational survey designs.

For a fixed survey length, the expected IBV is analytically available. However, for the sequential approaches this is not the case. For comparison the properties are therefore evaluated using Monte Carlo sampling over several replicates of realizations from the prior, conducting simulated sequential surveying for each one. Fig. \ref{fig:Eprob} shows the conditional EP, given data gathered along the north-south survey line for the realization indicated in these displays. In the Monte Carlo replicates, such results are repeatedly computed to approximate the expected IBV. We also compare predictive performance measured by root mean square error (RMSE) for the temperature and salinity estimates and the variance reduction in these two variables. It is important to note that the objective function used by the agents is focused on reducing the expected IBV, but we nevertheless expect that we will achieve good predictive performance for criteria such as RMSE as well. Another non-statistical criteria that is important for practical purposes is the computational time of the strategy, as this will impact the performance of the approach in the field for an embedded system.

@@

Each strategy is conducted on an equilateral grid as shown in Fig. \ref{fig:wp_graph}. The sequential sampling agent starts at the center East-West coordinate at the southern end of the domain (node 53). The AUV will move along edges in the waypoint graph while the vehicle gathers measurements. The data are assimilated into the GP model before an evaluation of the next node to sample is conducted at the end of the edge.

\subsection{Simulation Results}
A total of 100 replicate simulations were conducted. For each simulation all strategies are run. The results are shown in Fig. \ref{fig:sim_results},
where the different criteria are plotted as a function of survey distance. Fig. \ref{fig:avg_ev} shows the resulting drop in IBV for each of the six strategies. The IBV is reduced most under the \textit{myopic} and \textit{look-ahead} strategies, each performing almost equally; this is expected as the two criteria (Eq. \eqref{critSEQ} and \eqref{critLA}) are much more sensitive to differences in IBV. The \textit{static\_north} design also does well here because the path is parallel to the  boundary between the water masses.

@@

Fig. \ref{fig:avg_rmse} and \ref{fig:avg_r2} show the resulting drop in RMSE and increase in explained variance, respectively. Both \textit{myopic} and \textit{look-ahead} strategies perform well here, but some of the \textit{static\_east} and \textit{static\_zigzag} also achieve good results because they are pre-determined to cover large parts of the domain without re-visitation. Sequential strategies targeting IBV will sometimes not reach similar coverage, as interesting data may draw the path into twists and turns. There is relatively large variety in the replicate results as indicated by the vertical lines. Still, the ordering of strategies is similar.

@@

Fig.  \ref{fig:avg_time} shows the computational effort: the \textit{naive} strategy is on par with the static designs, while the \textit{myopic} strategy is slower. The \textit{look-ahead} is even slower, reaching levels that are nearly impractical for execution on a vehicle. Some pruning of the graph is performed to improve the computational speed, such as ruling out repeated visitations and back-and-forth routes. Some of the intermediate results are also stored for longer planning horizons. Still, more pruning of branches or inclusion of other heuristics could be included to make the strategy faster. Then again, the inclusion of heuristics is likely a contributing factor to the \textit{look-ahead} strategy failing to outperform the \textit{myopic} strategy.

In Fig. \ref{fig:route_choices}, the realized sampling paths for each of the sequential schemes and static designs are shown. The \textit{naive} strategy often gets stuck in the southern part of the domain because it is too focused on the probabilities near $0.5$. The \textit{myopic} strategy covers a wider domain than the naive or look-ahead. There are several reasons for this; first, a greedy approach will tend to put more emphasis on locations close to the agent, which may lead away from the centre. Second, as the agent evaluates the impact of locations further away (look-ahead) where assimilated data has less predictive power, the GP prior model (which is centered here) will act to restrict paths deviating from the central zone.

@@

\subsection{Performance Metrics}
The tables below explore the sensitivity of the result to different prior correlation values between temperature measured onboard the AUV, temperature and salinity, prior standard deviations, and spatial correlations (see table rows). We only compare results at the end of each survey. The expected ES variances are shown in Table \ref{tab:sim_res_ev}, results of RMSE are in Table \ref{tab:sim_res_rmse}, and explained variance is in Table \ref{tab:sim_res_r2}. In all runs, both the \textit{myopic} and \textit{look-ahead} strategies perform the best in terms of expected variance in the ES. Surprisingly, the \textit{myopic} performs slightly better than the \textit{look-ahead} strategy. Part of the reason for this performance discrepancy can be traced to the act of pruning some of the paths for the \textit{look-ahead} strategy; however, even with less pruning, the \textit{look-ahead} does not outperform the \textit{myopic} approach. The \textit{static\_north} design is clearly better than the other two static designs, the reason being that the others do not focus on the areas where the ES is really uncertain (centre locations are the most uncertain). By measuring temperature alone, the expected variances in the ES are only slightly higher than the combined temperature-salinity estimate. This occurs because the correlation of $0.6$ means that temperature observations provide information about salinity as well. For the other predictive measures, \textit{myopic} and \textit{look-ahead} strategies do well, clearly beating the \textit{naive} strategy. The \textit{static\_zigzag} design is usually the best among the static designs for these predictive performance measures because its pattern gives more effective coverage of the domain when there is significant spatial correlation, as indicated in this case.

@@

To demonstrate the applicability of using multivariate EPs and the IBV to inform oceanographic sampling, we present a case study mapping a river plume with an AUV. The experiment was performed in Trondheim, Norway,
surveying the Nidelva river (Fig. \ref{fig:nidelven}). The experiments were conducted in late Spring 2019, when there is still snow melting in the surrounding mountains so that the river water is substantially colder than the water in the fjord. The experiment is focused along the frontal zone that runs more or less parallel to the eastern shore as
indicated in purple in Fig. \ref{fig:nidelven}.

@@

The statistical model parameters were specified based on a short preliminary survey where the AUV did an initial transect to determine the environmental conditions and correlation structure. Based on the initial data, the trend parameters were estimated by linear regression, where both temperature and salinity are assumed to increase linearly, going west from the river mouth. Next, the residuals from the regression analysis were analyzed to study the fit of the GP model and to specify the covariance parameters.

@@

Fig. \ref{fig:parest} shows diagnostic plots of this analysis. Fig. \ref{fig:parest_a} shows a cross-plot of temperature and salinity residuals after the westernly trends in both salinity and temperature are subtracted from the data. This scatterplot of joint residuals indicates larger variability in salinity than in temperature, and a positive correlation ($0.5$) between the two variables. Based on the fitted bivariate Gaussian model (ellipses in Fig. \ref{fig:parest_a}), we can compute the modeled quadratic form of the residuals, and if the model is adequate they should be approximately $\chi^2_2$ distributed. Fig. \ref{fig:parest_b} shows the empirical cumulative distribution function (CDF) of these quadratic forms (solid) together with the theoretical CDF of the $\chi^2_2$ distribution. The modeled and theoretical curves are very similar, which indicates that the Gaussian model fits reasonably well. Even though there appears to be some clustering in both Fig. \ref{fig:parest_a} and \ref{fig:parest_b}, the bivariate diagnostic plots look reasonable and justify a Gaussian model. Fig. \ref{fig:parest_c} shows the empirical variogram of the scaled residuals for temperature and salinity. The decay is similar for the two, and seems to be negligible after about $150$ m.

@@

\begin{figure}[!h] 
\centering 
\includegraphics[width=0.98\textwidth]{Figures/harald.jpg}
\caption{The commercially available Light Autonomous Underwater
  Vehicle (LAUV) platform for upper water-column exploration used in
  the experiment.}
\label{fig:lauv}
\end{figure}

@@

The AUV was set to start in the south-center part of the waypoint graph, with the previously outlined GP model of the environment. A survey was set to take approximately 40 minutes to visit 15 waypoints on the grid, with the AUV running near the surface to better capture the plume. On its path from one waypoint to the next, the AUV gathered data regularly, and the GP model assimilated temperature and salinity data with an update frequency of 30 seconds, with about three updates per stage.

@@

Both Surveys 1 and 2 successfully estimated and navigated the separation zone, crossing the frontal boundary multiple times. As conditions changed slightly between the two deployments, the resulting path (after waypoint 5) deviates. Survey 1 continued northwards, tracking the north-eastern part of the front, while Survey 2 turned west, mapping the south-western part.

The final prediction of the front location given in Fig. \ref{fig:res1} and \ref{fig:res2} as a dashed line, correspond with one another, yielding a picture of the front gradually bending off toward the north east. The amount of exploration done by Survey 1 is greater than Survey 2. In Survey 1, the AUV obtained more detail by going north, while Survey 2, coming close to the survey area borders in the south-western corner, obtained a poorer understanding of the northern parts. A longer planning horizon would likely identify and discourage such choices.

@@

This paper builds on multidisciplinary work combining statistical modeling and methods with robotic surveying technologies for oceanographic sensing applications. We show how  observation practices can gain efficiency and accuracy from the development of statistical techniques to achieve more effective spatial monitoring. We further demonstrate the opportunities available for real-time multivariable spatial data-gathering and analysis onboard autonomous sensing platforms, which statisticians can exploit to create general-purpose toolkits for similar applications.

@@

The spatial-statistical design criterion, building on ESs, is relevant in our setting with different water mass properties; this could also be useful in other oceanographic settings, such as algae-blooms or frontal systems. Of course, other criteria are also relevant for decision-making.  For instance, managers and regulators must make difficult decisions related to fish farming, other marine resources operations, or environmental projects. Value of information analysis (VOI) \citep{Eidsvik:15} could be used in a similar vein as the IBV in the current paper, as both analytic methods explore when, where and which information is likely to result in improved decision-making.

While more effort could be spent on approximating the look-ahead steps in the adaptive designs, it is perhaps more interesting to explore the additional flexibility that can be gained by having two or more AUVs  exploring a spatial or spatio-temporal domain together \citep{ferreira2019advancing}. Such an approach would enable concurrent sampling in  different parts of the space, or opportunities to move in parallel to best capture the excursion set.

\section*{Acknowledgements}
TOF and KR acknowledge support from the Centre for Autonomous Marine Operations and Systems (AMOS)\footnote{\url{https://www.ntnu.edu/amos}}, Center of Excellence, project number 223254, and the Applied Underwater Robotics Labortatory (AURLab). JE acknowledges support from Norwegian research council project 294404. DG acknowledges support from the Swiss National Science Foundation, project number 178858.

>>>>

The covariance is assumed to be stationary and separable for the two variables. 
The $2 \times 2$ pointwise covariance matrix is set to:
\begin{equation}\label{v0}
\bSigma(\bx)=\mbox{Var} 
\begin{bmatrix}
    \xi_T(\bx) \\
    \xi_S(\bx) 
    \end{bmatrix}=
\begin{bmatrix}
0.25^2 & 0.6 \cdot 0.25^2 \\
0.6 \cdot 0.25^2 & 0.25^2
\end{bmatrix}.
\end{equation}
The spatial correlation is of a Matern type \kc{cite or define}:
\begin{equation}\label{v}
\small
\mbox{Corr} 
\left(
\begin{bmatrix}
    \xi_T(\bx) \\
    \xi_S(\bx) 
    \end{bmatrix},
    \begin{bmatrix}
    \xi_T(\bx') \\
    \xi_S(\bx') 
    \end{bmatrix}
    \right)
    = \begin{bmatrix}
1 & 0.6  \\
0.6  & 1
\end{bmatrix}(1+\phi h)\exp (-\phi h),
\end{equation}
where $h=\sqrt{||\bx-\bx'||^2}$ and $\phi=0.3$ indicates an effective correlation range of about $1200$ m. 
%Figure \ref{fig:stat_design} shows the contour lines for the EP for the reference. We notice the trend of increasing salinity and temperature to the west.  
One realization of the random fields representing salinity and
temperature is shown in Fig. \ref{fig:true_temp} and
\ref{fig:true_sal}. The true ES from this prior realization is shown
in Fig. \ref{fig:ESet}.

@@

We compare three different static designs denoted
\textit{static\_north}, \textit{static\_east}, and
\textit{static\_zigzag} with the three sequential approaches
\textit{naive}, \textit{myopic}, and \textit{look-ahead}. Recall that
the static sampling paths are pre-determined and cannot be altered.
These represent the pre-planned strategies used in most current AUV
operational survey designs.

For a fixed survey length, the expected IBV is analytically available
\kc{I don't understand what ``analytically available''
  means}. However, for the sequential approaches this is not the
case. For comparison the properties are therefore evaluated using
Monte Carlo sampling over several replicates of realizations from the
prior while conducting simulated sequential surveys for each
one. Fig. \ref{fig:Eprob} shows the conditional EP, given data
gathered along the north-south survey lines in these figures.  In the
Monte Carlo replicates, such results are repeatedly computed to
approximate the expected IBV. We also compare predictive performance
measured by root mean square error (RMSE) for temperature and salinity
estimates as also the variance reduction in these two variables. It is
important to note that the objective function used by the agent is
focused on reducing the expected IBV, but we nevertheless expect that
we will achieve good predictive performance for criteria such as RMSE
as well. Another non-statistical criteria that is important for
practical purposes is the computational time needed for the strategy,
as this will impact the performance for an embedded system.

@@

Each strategy is conducted on an equilateral grid as shown in
Fig. \ref{fig:wp_graph}. The sequential sampling agent starts at the
center East-West coordinate at the southern end of the domain (node
53). The AUV moves along edges in the waypoint graph while the vehicle
gathers measurements. The data is assimilated into the GP model before
an evaluation of the next node to sample is conducted at the end of
the edge. \kc{Note there are plenty of language errors related to
  tense. Pl. do not mess with any more changes from here onwards.}

\subsection{Simulation Results}

A total of 100 replicate simulations were conducted with all
strategies. The results are shown in Fig. \ref{fig:sim_results}, where
the different criteria are plotted as a function of survey
distance. Fig. \ref{fig:avg_ev} shows the resulting drop in IBV for
each of the six strategies. IBV reduction occurs most under the
\textit{myopic} and \textit{look-ahead} strategies, each performing
almost equally; this is expected as the two criteria
(Eq. \eqref{critSEQ} and \eqref{critLA}) are sensitive to differences
in IBV. The \textit{static\_north} design also does well here because
the path is parallel to the boundary between the water masses.

@@

Fig. \ref{fig:avg_rmse} and \ref{fig:avg_r2} show the resulting drop
in RMSE and increase in explained variance, respectively. Both
\textit{myopic} and \textit{look-ahead} strategies perform well here,
but some of the \textit{static\_east} and \textit{static\_zigzag} also
achieve good results because they are pre-determined to cover large
parts of the domain without re-visitation. Sequential strategies
targeting IBV will sometimes not reach similar coverage, as
interesting data may draw the path into twists and turns. There is
relatively large variety in the replicate results as indicated by the
vertical lines. Nevertheless, the ordering of strategies is similar.

@@

Fig.  \ref{fig:avg_time} shows the computational effort: the
\textit{naive} strategy is on par with the static designs, while the
\textit{myopic} strategy is slower. The \textit{look-ahead} is even
slower, reaching levels that are nearly impractical for execution on a
vehicle. Some pruning of the graph is performed to improve the
computational speed, such as ruling out repeated visitations and
back-and-forth routes. Some of the intermediate results are also
stored for longer planning horizons. Further pruning of branches or
inclusion of other heuristics could be included to make the strategy
faster. Then again, the inclusion of heuristics is likely a
contributing factor for the \textit{look-ahead} strategy failing to
outperform the \textit{myopic} strategy.

In Fig. \ref{fig:route_choices}, the realized sampling paths for each
of the sequential schemes and static designs are shown. The
\textit{naive} strategy often gets stuck in the southern part of the
domain because it is too focused on the probabilities near $0.5$. The
\textit{myopic} strategy covers a wider domain than the naive or
look-ahead. There are several reasons for this. First, a greedy
approach will tend to put more emphasis on locations close to the
agent, which may lead away from the centre. Second, as the agent
evaluates the impact of locations further away (look-ahead) where
assimilated data has less predictive power, the GP prior model (which
is centered here) will act to restrict paths deviating from the
central zone.

@@

\subsection{Performance Metrics}

The tables below explore the sensitivity of the result to different
prior correlation values between temperature measured onboard the AUV,
temperature and salinity, prior standard deviations, and spatial
correlations (see table rows). We only compare results at the end of
each survey. The expected ES variances are shown in Table
\ref{tab:sim_res_ev}, results of RMSE are in Table
\ref{tab:sim_res_rmse}, and explained variance is in Table
\ref{tab:sim_res_r2}. In all runs, both the \textit{myopic} and
\textit{look-ahead} strategies perform the best in terms of expected
variance in the ES. Surprisingly, the \textit{myopic} strategy
performs slightly better than the \textit{look-ahead}. Part of the
reason for this discrepancy can be traced to the act of pruning some
of the paths for the \textit{look-ahead} strategy; however, even with
less pruning, the \textit{look-ahead} does not outperform the
\textit{myopic} approach. The \textit{static\_north} design is clearly
better than the other two static designs, the reason being that the
others do not focus on the areas where the ES is really uncertain
(centre locations are the most uncertain). By measuring temperature
alone, the expected variances in the ES are only slightly higher than
the combined temperature-salinity estimate. This occurs because the
correlation of $0.6$ means that temperature observations provide
information about salinity as well. For the other predictive measures,
\textit{myopic} and \textit{look-ahead} strategies do well, clearly
beating the \textit{naive} strategy. The \textit{static\_zigzag}
design is usually the best among the static designs for these
predictive performance measures because its pattern gives more
effective coverage of the domain when there is significant spatial
correlation, as indicated in this case.

@@

To demonstrate the applicability of using multivariate EPs and the IBV
to inform oceanographic sampling, we present a case study mapping a
river plume with an AUV. The experiment was performed in Trondheim,
Norway, surveying the Nidelva river (Fig. \ref{fig:nidelven}). The
experiments were conducted in late Spring 2019, when there is still
snow melting in the surrounding mountains so that the river water is
substantially colder than the water in the fjord. The experiment was
focused along the frontal zone that runs more or less parallel to the
eastern shore as noted in Fig. \ref{fig:nidelven}.

@@

The statistical model parameters were specified based on a short
preliminary survey where the AUV made an initial transect to determine
environmental conditions and correlation structure. Based on the
initial data, the trend parameters were estimated by linear
regression, where both temperature and salinity are assumed to
increase linearly, going west from the river mouth. Next, the
residuals from the regression analysis were analyzed to study the fit
of the GP model and to specify the covariance parameters.

@@

Fig. \ref{fig:parest} shows diagnostic plots of this
analysis. Fig. \ref{fig:parest_a} shows a cross-plot of temperature
and salinity residuals after the westerly trends in both salinity and
temperature are subtracted from the data. This scatterplot of joint
residuals indicates larger variability in salinity than in
temperature, and a positive correlation ($0.5$) between the two
variables. Based on the fitted bivariate Gaussian model (ellipses in
Fig. \ref{fig:parest_a}), we can compute the modeled quadratic form of
the residuals, and if the model is adequate they should be
approximately $\chi^2_2$ distributed. Fig. \ref{fig:parest_b} shows
the empirical cumulative distribution function (CDF) of these
quadratic forms (solid) together with the theoretical CDF of the
$\chi^2_2$ distribution. The modeled and theoretical curves are very
similar, which indicates that the Gaussian model fits reasonably
well. Even though there appears to be some clustering in both
Fig. \ref{fig:parest_a} and \ref{fig:parest_b}, the bivariate
diagnostic plots look reasonable and justify a Gaussian
model. Fig. \ref{fig:parest_c} shows the empirical variogram of the
scaled residuals for temperature and salinity. The decay is similar
for the two, and seems to be negligible after about $150$ m.

@@

\begin{figure}[!h] 
\centering 
\includegraphics[width=0.98\textwidth]{Figures/harald.jpg}
\caption{The commercially available Light Autonomous Underwater
  Vehicle (LAUV) platform for upper water-column exploration used in
  our experiments.}
\label{fig:lauv}
\end{figure}

@@

The AUV was set to start in the south-center part of the waypoint
graph, with the previously outlined GP model of the environment. A
survey was set to take approximately 40 minutes to visit 15 waypoints
on the grid, with the AUV running near the surface to capture the
plume. On its path from one waypoint to the next, the AUV gathered
data regularly, and the GP model assimilated temperature and salinity
data with an update frequency of 30 seconds, with about three updates
per stage.

@@

Both surveys successfully estimated and navigated the separation zone,
crossing the frontal boundary multiple times. As conditions changed
slightly between the two deployments, the resulting path (after
waypoint 5) is shown to deviate. Survey 1 continued northwards,
tracking the north-eastern part of the front, while Survey 2 turned
west, mapping the south-western part.

The final predictions of the front location given in
Fig. \ref{fig:res1} and \ref{fig:res2} as a dashed line, correspond
with one another, yielding a picture of the front gradually bending
off toward the north east. The amount of exploration done by Survey 1
is greater than Survey 2. In Survey 1, the AUV obtained more detail by
going north, while Survey 2, coming close to the survey area borders
in the south-western corner, obtained a poorer understanding of the
northern parts. A longer planning horizon would likely identify and
discourage such choices.

@@

This work builds on a multidisciplinary effort combining statistical
modeling with robotic surveying techniques for oceanographic
applications. We show how observation practices can gain efficiency
and accuracy from the development of statistical techniques to achieve
more effective spatial monitoring. We further demonstrate the
opportunities available for real-time multivariable spatial
data-gathering and analysis onboard autonomous sensing platforms,
which statisticians can exploit to create general-purpose toolkits for
similar applications.

@@

The spatial-statistical design criterion, building on ESs, is relevant
in our setting with different water mass properties; this could also
be useful in other oceanographic settings, such as algal-blooms or
open water fronts. Of course, other criteria are also relevant for
decision-making.  For instance, managers and regulators must make
difficult decisions related to fish farming, other marine resources
operations, or environmental projects. Value of information analysis
(VOI) \citep{Eidsvik:15} could be used in a similar vein as the IBV in
the current paper, as both analytic methods explore when, where and
which information is likely to result in improved decision-making.

While more effort could be spent on approximating the look-ahead steps
in the adaptive designs, it is perhaps more interesting to explore the
additional flexibility that can be gained by having two or more AUVs
co-temporally exploring a spatial or spatio-temporal domain together
\citep{ferreira2019advancing}. Such an approach would enable
concurrent sampling in different parts of the space, or opportunities
to move in parallel to best capture the excursion set.

\section*{Acknowledgements}

TOF and KR acknowledge support from the Centre for Autonomous Marine
Operations and Systems
(AMOS)\footnote{\url{https://www.ntnu.edu/amos}}, Center of
Excellence, project number 223254, and the Applied Underwater Robotics
Labortatory (AURLab). JE acknowledges support from Norwegian research
council project 294404. DG acknowledges support from the Swiss
National Science Foundation, project number 178858.
